---
title: "Divergence analysis"
author: "James Ashmore"
date: "22/02/2021"
output: html_document
---

## Setup

Set chunk options:

```{r}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = FALSE,
  dev        = "png",
  echo       = TRUE,
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE
)
```

Load Bioconductor packages:

```{r}
library(scater)
library(scran)
library(scuttle)
```

Load CRAN packages:

```{r}
library(ggforce)
library(here)
```

Define helper functions:

```{r}
resizeGuides <- function(object, pointSize = 0.5, textSize = 3, spaceLegend = 0.1) {
    object + 
        guides(
            shape = guide_legend(ncol = 1, override.aes = list(size = pointSize)),
            color = guide_legend(ncol = 1, override.aes = list(size = pointSize))
        ) + 
        theme(
            legend.title = element_text(size = textSize),
            legend.text  = element_text(size = textSize),
            legend.key.size = unit(spaceLegend, "lines")
        )
}
```

Define colour palettes:

```{r}
EmbryoStageColours <- c(
    "E6.5"               = "#D53E4F",
    "E6.75"              = "#F46D43",
    "E7.0"               = "#FDAE61",
    "E7.25"              = "#FEE08B",
    "E7.5"               = "#FFFFBF",
    "E7.75"              = "#E6F598",
    "E8.0"               = "#ABDDA4",
    "E8.25"              = "#66C2A5",
    "E8.5"               = "#3288BD",
    "mixed_gastrulation" = "#A9A9A9"
)

EmbryoTheilerColours <- c(
    "TS9"    = "#D7191C",
    "TS10"   = "#FDAE61",
    "TS11"   = "#ABDDA4",
    "TS12"   = "#2B83BA",
    "TS9-10" = "#A9A9A9"
)

EmbryoCelltypeColours <- c(
    "Allantois"                      = "#532C8A",
    "Anterior Primitive Streak"      = "#C19F70",
    "Blood progenitors 1"            = "#F9DECF",
    "Blood progenitors 2"            = "#C9A997",
    "Cardiomyocytes"                 = "#B51D8D",
    "Caudal Mesoderm"                = "#3F84AA",
    "Caudal epiblast"                = "#9E6762",
    "Caudal neurectoderm"            = "#354E23",
    "Def. endoderm"                  = "#F397C0",
    "Endothelium"                    = "#FF891C",
    "Epiblast"                       = "#635547",
    "Erythroid1"                     = "#C72228",
    "Erythroid2"                     = "#F79083",
    "Erythroid3"                     = "#EF4E22",
    "ExE ectoderm"                   = "#989898",
    "ExE endoderm"                   = "#7F6874",
    "ExE mesoderm"                   = "#8870AD",
    "Forebrain/Midbrain/Hindbrain"   = "#647A4F",
    "Gut"                            = "#EF5A9D",
    "Haematoendothelial progenitors" = "#FBBE92",
    "Intermediate mesoderm"          = "#139992",
    "Mesenchyme"                     = "#CC7818",
    "Mixed mesoderm"                 = "#DFCDE4",
    "NMP"                            = "#8EC792",
    "Nascent mesoderm"               = "#C594BF",
    "Neural crest"                   = "#C3C388",
    "Notochord"                      = "#0F4A9C",
    "PGC"                            = "#FACB12",
    "Paraxial mesoderm"              = "#8DB5CE",
    "Parietal endoderm"              = "#1A1A1A",
    "Pharyngeal mesoderm"            = "#C9EBFB",
    "Primitive Streak"               = "#DABE99",
    "Rostral neurectoderm"           = "#65A83E",
    "Somitic mesoderm"               = "#005579",
    "Spinal cord"                    = "#CDE088",
    "Surface ectoderm"               = "#F7F79E",
    "Visceral endoderm"              = "#F6BFCB"
)
```

Read experiment object:

```{r}
con <- here("data/EmbryoAtlasData.LPMP.rds")

sce <- readRDS(con)
```

Define source annotation:

```{r}
sce$source <- "MGD"

sce$source[sce$sample == "EHF"] <- "EHF"

sce$source[sce$sample == "LHF"] <- "LHF"
```

Define data annotation:

```{r}
sce$data <- "reference"

sce$data[sce$sample == "EHF"] <- "query"

sce$data[sce$sample == "LHF"] <- "query"
```

## Exploration

Create a per-cell data frame:

```{r}
dat <- makePerCellDF(sce, use.coldata = TRUE, use.dimred = "UMAP")

dat <- split(dat, dat$data)
```

Plot cells coloured by stage annotation:

```{r}
ggplot(dat$reference, aes(UMAP.1, UMAP.2, colour = stage)) + 
    geom_point(size = 0.1) + 
    scale_colour_manual(name = "Stage", values = EmbryoStageColours) + 
    guides(colour = guide_legend(override.aes = list(size = 2))) + 
    geom_point(data = dat$query, aes(UMAP.1, UMAP.2, shape = sample), inherit.aes = FALSE) + 
    scale_shape_manual(name = "Sample", values = c(1, 4)) + 
    theme_no_axes() + 
    theme(aspect.ratio = 1)
```

Plot cells coloured by theiler annotation:

```{r}
ggplot(dat$reference, aes(UMAP.1, UMAP.2, colour = theiler)) + 
    geom_point(size = 0.1) + 
    scale_colour_manual(name = "Theiler", values = EmbryoTheilerColours, na.translate = FALSE) + 
    guides(colour = guide_legend(override.aes = list(size = 2))) + 
    geom_point(data = dat$query, aes(UMAP.1, UMAP.2, shape = sample), inherit.aes = FALSE) + 
    scale_shape_manual(name = "Sample", values = c(1, 4)) + 
    theme_no_axes() + 
    theme(aspect.ratio = 1)
```

Plot cells coloured by celltype annotation:

```{r}
plt <- ggplot(dat$reference, aes(UMAP.1, UMAP.2, colour = celltype)) + 
    geom_point(size = 0.1) + 
    scale_colour_manual(name = "Celltype", values = EmbryoCelltypeColours, na.translate = FALSE) + 
    geom_point(data = dat$query, aes(UMAP.1, UMAP.2, shape = sample), inherit.aes = FALSE) + 
    scale_shape_manual(name = "Sample", values = c(1, 4)) + 
    theme_no_axes() + 
    theme(aspect.ratio = 1)

resizeGuides(plt, pointSize = 1, textSize = 7, spaceLegend = 0.6)
```

## Divergence {.tabset}

Create a per-cell data frame:

```{r}
dat <- makePerCellDF(sce, use.coldata = TRUE, use.dimred = "UMAP")
```

Define colour annotation:

```{r}
colour.values <- c("MGD" = "transparent", "EHF" = "#E15759", "LHF" = "#4E79A7")

colour.breaks <- c("EHF", "LHF")
```

### Endothelium vs Paraxial mesoderm

Plot cells coloured by Endothelium odds annotation:

```{r}
dat$odds_8.5_endothelium <- dat$fates_8.5_endothelium / (1 - dat$fates_8.5_endothelium)

ggplot(dat, aes(UMAP.1, UMAP.2, colour = source, fill = odds_8.5_endothelium)) +
    geom_point(shape = 21) +
    scale_fill_gradient(name = "Odds (Endothelium)", low = "#EBEBEB", high = EmbryoCelltypeColours["Endothelium"]) + 
    scale_colour_manual(name = "Source", values = colour.values, breaks = colour.breaks) + 
    facet_wrap( ~ stage, nrow = 2) +
    theme_no_axes() +
    theme(aspect.ratio = 1, legend.position = "top")
```

Plot cells coloured by Paraxial mesoderm odds annotation:

```{r}
dat$odds_8.5_paraxial.mesoderm <- dat$fates_8.5_paraxial.mesoderm / (1 - dat$fates_8.5_paraxial.mesoderm)

ggplot(dat, aes(UMAP.1, UMAP.2, colour = source, fill = odds_8.5_paraxial.mesoderm)) +
    geom_point(shape = 21) +
    scale_fill_gradient(name = "Odds (Paraxial mesoderm)", low = "#EBEBEB", high = EmbryoCelltypeColours["Paraxial mesoderm"]) +
    scale_colour_manual(name = "Source", values = colour.values, breaks = colour.breaks) + 
    facet_wrap( ~ stage, nrow = 2) +
    theme_no_axes() +
    theme(aspect.ratio = 1, legend.position = "top")
```

Plot cells coloured by Endothelium vs Paraxial mesoderm odds annotation:

```{r}
dat$odds_8.5_endothelium_paraxial.mesoderm <- log(dat$odds_8.5_endothelium / dat$odds_8.5_paraxial.mesoderm)

ggplot(dat, aes(UMAP.1, UMAP.2, colour = source, fill = odds_8.5_endothelium_paraxial.mesoderm)) +
    geom_point(shape = 21) +
    scale_fill_gradient2(
        name = "Log Odds (Endothelium / Paraxial mesoderm)",
        high = EmbryoCelltypeColours["Endothelium"],
        mid = "#EBEBEB",
        low = EmbryoCelltypeColours["Paraxial mesoderm"]
    ) +
    scale_colour_manual(name = "Source", values = colour.values, breaks = colour.breaks) + 
    facet_wrap( ~ stage, nrow = 2) +
    theme_no_axes() +
    theme(aspect.ratio = 1, legend.position = "top")
```

### Allantois vs Cardiomyocytes

Plot cells coloured by Allantois odds annotation:

```{r}
dat$odds_8.5_allantois <- dat$fates_8.5_allantois / (1 - dat$fates_8.5_allantois)

ggplot(dat, aes(UMAP.1, UMAP.2, colour = source, fill = odds_8.5_allantois)) +
    geom_point(shape = 21) +
    scale_fill_gradient(name = "Odds (Allantois)", low = "#EBEBEB", high = EmbryoCelltypeColours["Allantois"]) +
    scale_colour_manual(name = "Source", values = colour.values, breaks = colour.breaks) + 
    facet_wrap( ~ stage, nrow = 2) +
    theme_no_axes() +
    theme(aspect.ratio = 1, legend.position = "top")
```

Plot cells coloured by Cardiomyocytes odds annotation:

```{r}
dat$odds_8.5_cardiomyocytes <- dat$fates_8.5_cardiomyocytes / (1 - dat$fates_8.5_cardiomyocytes)

ggplot(dat, aes(UMAP.1, UMAP.2, colour = source, fill = odds_8.5_cardiomyocytes)) +
    geom_point(shape = 21) +
    scale_fill_gradient(name = "Odds (Cardiomyocytes)", low = "#EBEBEB", high = EmbryoCelltypeColours["Cardiomyocytes"]) +
    scale_colour_manual(name = "Source", values = colour.values, breaks = colour.breaks) + 
    facet_wrap( ~ stage, nrow = 2) +
    theme_no_axes() +
    theme(aspect.ratio = 1, legend.position = "top")
```

Plot cells coloured by Allantois vs Cardiomyocytes odds annotation:

```{r}
dat$odds_8.5_allantois_cardiomyocytes <- log(dat$odds_8.5_allantois / dat$odds_8.5_cardiomyocytes)

ggplot(dat, aes(UMAP.1, UMAP.2, colour = source, fill = odds_8.5_allantois_cardiomyocytes)) +
    geom_point(shape = 21) +
    scale_fill_gradient2(
        name = "Log Odds (Allantois / Cardiomyocytes)",
        high = EmbryoCelltypeColours["Allantois"],
        mid = "#EBEBEB",
        low = EmbryoCelltypeColours["Cardiomyocytes"]
    ) +
    scale_colour_manual(name = "Source", values = colour.values, breaks = colour.breaks) + 
    facet_wrap( ~ stage, nrow = 2) +
    theme_no_axes() +
    theme(aspect.ratio = 1, legend.position = "top")
```

### Cardiomyocytes vs Pharyngeal mesoderm

Plot cells coloured by Cardiomyocytes odds annotation:

```{r}
dat$odds_8.5_cardiomyocytes <- dat$fates_8.5_cardiomyocytes / (1 - dat$fates_8.5_cardiomyocytes)

ggplot(dat, aes(UMAP.1, UMAP.2, colour = source, fill = odds_8.5_cardiomyocytes)) +
    geom_point(shape = 21) +
    scale_fill_gradient(name = "Odds (Cardiomyocytes)", low = "#EBEBEB", high = EmbryoCelltypeColours["Cardiomyocytes"]) +
    scale_colour_manual(name = "Source", values = colour.values, breaks = colour.breaks) + 
    facet_wrap( ~ stage, nrow = 2) +
    theme_no_axes() +
    theme(aspect.ratio = 1, legend.position = "top")
```

Plot cells coloured by Pharyngeal mesoderm odds annotation:

```{r}
dat$odds_8.5_pharyngeal.mesoderm <- dat$fates_8.5_pharyngeal.mesoderm / (1 - dat$fates_8.5_pharyngeal.mesoderm)

ggplot(dat, aes(UMAP.1, UMAP.2, colour = source, fill = odds_8.5_pharyngeal.mesoderm)) +
    geom_point(shape = 21) +
    scale_fill_gradient(name = "Odds (Pharyngeal mesoderm)", low = "#EBEBEB", high = EmbryoCelltypeColours["Pharyngeal mesoderm"]) +
    scale_colour_manual(name = "Source", values = colour.values, breaks = colour.breaks) + 
    facet_wrap( ~ stage, nrow = 2) +
    theme_no_axes() +
    theme(aspect.ratio = 1, legend.position = "top")
```

Plot cells coloured by Cardiomyocytes vs Pharyngeal mesoderm odds annotation:

```{r}
dat$odds_8.5_cardiomyocytes_pharyngeal.mesoderm <- log(dat$odds_8.5_cardiomyocytes / dat$odds_8.5_pharyngeal.mesoderm)

ggplot(dat, aes(UMAP.1, UMAP.2, colour = source, fill = odds_8.5_cardiomyocytes_pharyngeal.mesoderm)) +
    geom_point(shape = 21) +
    scale_fill_gradient2(
        name = "Log Odds (Cardiomyocytes / Pharyngeal mesoderm)",
        high = EmbryoCelltypeColours["Cardiomyocytes"],
        mid = "#EBEBEB",
        low = EmbryoCelltypeColours["Pharyngeal mesoderm"]
    ) +
    scale_colour_manual(name = "Source", values = colour.values, breaks = colour.breaks) + 
    facet_wrap( ~ stage, nrow = 2) +
    theme_no_axes() +
    theme(aspect.ratio = 1, legend.position = "top")
```

### ExE mesoderm vs Mesenchyme

Plot cells coloured by ExE mesoderm odds annotation:

```{r}
dat$odds_8.5_extraembryonic.mesoderm <- dat$fates_8.5_extraembryonic.mesoderm / (1 - dat$fates_8.5_extraembryonic.mesoderm)

ggplot(dat, aes(UMAP.1, UMAP.2, colour = source, fill = odds_8.5_extraembryonic.mesoderm)) +
    geom_point(shape = 21) +
    scale_fill_gradient(name = "Odds (ExE mesoderm)", low = "#EBEBEB", high = EmbryoCelltypeColours["ExE mesoderm"]) +
    scale_colour_manual(name = "Source", values = colour.values, breaks = colour.breaks) + 
    facet_wrap( ~ stage, nrow = 2) +
    theme_no_axes() +
    theme(aspect.ratio = 1, legend.position = "top")
```

Plot cells coloured by Mesenchyme odds annotation:

```{r}
dat$odds_8.5_mesenchyme <- dat$fates_8.5_mesenchyme / (1 - dat$fates_8.5_mesenchyme)

ggplot(dat, aes(UMAP.1, UMAP.2, colour = source, fill = odds_8.5_mesenchyme)) +
    geom_point(shape = 21) +
    scale_fill_gradient(name = "Odds (Mesenchyme)", low = "#EBEBEB", high = EmbryoCelltypeColours["Mesenchyme"]) +
    scale_colour_manual(name = "Source", values = colour.values, breaks = colour.breaks) + 
    facet_wrap( ~ stage, nrow = 2) +
    theme_no_axes() +
    theme(aspect.ratio = 1, legend.position = "top")
```

Plot cells coloured by ExE mesoderm vs Mesenchyme odds annotation:

```{r}
dat$odds_8.5_extraembryonic.mesoderm_mesenchyme <- log(dat$odds_8.5_extraembryonic.mesoderm / dat$odds_8.5_mesenchyme)

ggplot(dat, aes(UMAP.1, UMAP.2, colour = source, fill = odds_8.5_extraembryonic.mesoderm_mesenchyme)) +
    geom_point(shape = 21) +
    scale_fill_gradient2(
        name = "Log Odds (ExE mesoderm / Mesenchyme)",
        high = EmbryoCelltypeColours["ExE mesoderm"],
        mid = "#EBEBEB",
        low = EmbryoCelltypeColours["Mesenchyme"]
    ) +
    scale_colour_manual(name = "Source", values = colour.values, breaks = colour.breaks) + 
    facet_wrap( ~ stage, nrow = 2) +
    theme_no_axes() +
    theme(aspect.ratio = 1, legend.position = "top")
```

## Summary

### Session

Print session information:

```{r}
sessionInfo()
```
