---
title: "Lineage drivers"
author: "James Ashmore"
date: "2021-08-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

Set chunk options:

```{r}
knitr::opts_chunk$set(
  cache      = FALSE,
  dev        = "png",
  echo       = TRUE,
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE
)
```

```{r, include = FALSE}
DT::datatable(matrix()) # Import datatable options, quietly
```

Load Bioconductor packages:

```{r}
library(batchelor)
library(scater)
library(scran)
library(scuttle)
library(slingshot)
```

Load CRAN packages:

```{r}
library(ggforce)
library(patchwork)
```

Source helper functions:

```{r}
source("code/intersectRows.R")
source("code/quickCorrect.R")
source("code/dist_clusters_diag.R")
source("code/resizeGuides.R")
```

Define aesthetic labels:

```{r}
EmbryoFateLabels <- c(
  "fate_7.25_ehf" = "Fate/E7.25/EHF", 
  "fate_7.5_ehf"  = "Fate/E7.5/EHF", 
  "fate_7.75_ehf" = "Fate/E7.75/EHF", 
  "fate_8.0_ehf"  = "Fate/E8.0/EHF",
  "fate_7.5_lhf"  = "Fate/E7.5/LHF",
  "fate_7.75_lhf" = "Fate/E7.75/LHF",
  "fate_8.0_lhf"  = "Fate/E8.0/LHF"
)

EmbryoTrajectoryLabels <- c(
  "trajectory_7.25_ehf" = "Trajectory/E7.25/EHF", 
  "trajectory_7.5_ehf"  = "Trajectory/E7.5/EHF", 
  "trajectory_7.75_ehf" = "Trajectory/E7.75/EHF", 
  "trajectory_8.0_ehf"  = "Trajectory/E8.0/EHF",
  "trajectory_7.5_lhf"  = "Trajectory/E7.5/LHF",
  "trajectory_7.75_lhf" = "Trajectory/E7.75/LHF",
  "trajectory_8.0_lhf"  = "Trajectory/E8.0/LHF"
)
```

Define colour palettes:

```{r}
EmbryoStageColours <- c(
    "E6.5"               = "#D53E4F",
    "E6.75"              = "#F46D43",
    "E7.0"               = "#FDAE61",
    "E7.25"              = "#FEE08B",
    "E7.5"               = "#FFFFBF",
    "E7.75"              = "#E6F598",
    "E8.0"               = "#ABDDA4",
    "E8.25"              = "#66C2A5",
    "E8.5"               = "#3288BD",
    "mixed_gastrulation" = "#A9A9A9"
)

EmbryoTheilerColours <- c(
    "TS9"    = "#D7191C",
    "TS10"   = "#FDAE61",
    "TS11"   = "#ABDDA4",
    "TS12"   = "#2B83BA",
    "TS9-10" = "#A9A9A9"
)

EmbryoCelltypeColours <- c(
    "Allantois"                      = "#532C8A",
    "Anterior Primitive Streak"      = "#C19F70",
    "Blood progenitors 1"            = "#F9DECF",
    "Blood progenitors 2"            = "#C9A997",
    "Cardiomyocytes"                 = "#B51D8D",
    "Caudal Mesoderm"                = "#3F84AA",
    "Caudal epiblast"                = "#9E6762",
    "Caudal neurectoderm"            = "#354E23",
    "Def. endoderm"                  = "#F397C0",
    "Endothelium"                    = "#FF891C",
    "Epiblast"                       = "#635547",
    "Erythroid1"                     = "#C72228",
    "Erythroid2"                     = "#F79083",
    "Erythroid3"                     = "#EF4E22",
    "ExE ectoderm"                   = "#989898",
    "ExE endoderm"                   = "#7F6874",
    "ExE mesoderm"                   = "#8870AD",
    "Forebrain/Midbrain/Hindbrain"   = "#647A4F",
    "Gut"                            = "#EF5A9D",
    "Haematoendothelial progenitors" = "#FBBE92",
    "Intermediate mesoderm"          = "#139992",
    "Mesenchyme"                     = "#CC7818",
    "Mixed mesoderm"                 = "#DFCDE4",
    "NMP"                            = "#8EC792",
    "Nascent mesoderm"               = "#C594BF",
    "Neural crest"                   = "#C3C388",
    "Notochord"                      = "#0F4A9C",
    "PGC"                            = "#FACB12",
    "Paraxial mesoderm"              = "#8DB5CE",
    "Parietal endoderm"              = "#1A1A1A",
    "Pharyngeal mesoderm"            = "#C9EBFB",
    "Primitive Streak"               = "#DABE99",
    "Rostral neurectoderm"           = "#65A83E",
    "Somitic mesoderm"               = "#005579",
    "Spinal cord"                    = "#CDE088",
    "Surface ectoderm"               = "#F7F79E",
    "Visceral endoderm"              = "#F6BFCB"
)

EmbryoSourceColours <- c(
    "MGD" = "#CFCFCF",
    "EHF" = "#E15759",
    "LHF" = "#4E79A7"
)
```

Read experiment object:

```{r}
sce <- readRDS("data/EmbryoAtlasData.LPMP.rds")
```

Define source annotation:

```{r}
sce$source <- "MGD"

sce$source[sce$sample == "EHF"] <- "EHF"

sce$source[sce$sample == "LHF"] <- "LHF"
```

Define data annotation:

```{r}
sce$data <- "reference"

sce$data[sce$sample == "EHF"] <- "query"

sce$data[sce$sample == "LHF"] <- "query"
```

## Exploration

Create a per-cell data frame:

```{r}
dat <- makePerCellDF(sce)

dat <- split(dat, dat$data)
```

Plot cells coloured by stage annotation:

```{r}
ggplot(dat$reference, aes(UMAP.1, UMAP.2, colour = stage)) + 
    geom_point(size = 0.1) + 
    scale_colour_manual(name = "Stage", values = EmbryoStageColours) + 
    guides(colour = guide_legend(override.aes = list(size = 2))) + 
    geom_point(data = dat$query, aes(UMAP.1, UMAP.2, shape = sample), inherit.aes = FALSE) + 
    scale_shape_manual(name = "Sample", values = c(1, 4)) + 
    theme_no_axes() + 
    theme(aspect.ratio = 1)
```

Plot cells coloured by theiler annotation:

```{r}
ggplot(dat$reference, aes(UMAP.1, UMAP.2, colour = theiler)) + 
    geom_point(size = 0.1) + 
    scale_colour_manual(name = "Theiler", values = EmbryoTheilerColours) + 
    guides(colour = guide_legend(override.aes = list(size = 2))) + 
    geom_point(data = dat$query, aes(UMAP.1, UMAP.2, shape = sample), inherit.aes = FALSE) + 
    scale_shape_manual(name = "Sample", values = c(1, 4)) + 
    theme_no_axes() + 
    theme(aspect.ratio = 1)
```

Plot cells coloured by celltype annotation:

```{r}
plt <- ggplot(dat$reference, aes(UMAP.1, UMAP.2, colour = celltype)) + 
    geom_point(size = 0.1) + 
    scale_colour_manual(name = "Celltype", values = EmbryoCelltypeColours) + 
    geom_point(data = dat$query, aes(UMAP.1, UMAP.2, shape = sample), inherit.aes = FALSE) + 
    scale_shape_manual(name = "Sample", values = c(1, 4)) + 
    theme_no_axes() + 
    theme(aspect.ratio = 1)

resizeGuides(plt, pointSize = 1, textSize = 7, spaceLegend = 0.6)
```

## Trajectory

Create a per-cell data frame:

```{r}
dat <- makePerCellDF(sce)
```

Plot E7.25 EHF trajectory:

```{r}
ggplot(dat, aes(UMAP.1, UMAP.2, colour = trajectory_7.25_ehf)) + 
  geom_point(size = 0.1) + 
  scale_colour_viridis_c(name = "Probability") + 
  ggtitle("Trajectory/E7.25/EHF") + 
  facet_grid(source ~ stage) + 
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

Plot E7.5 EHF trajectory:

```{r}
ggplot(dat, aes(UMAP.1, UMAP.2, colour = trajectory_7.5_ehf)) + 
  geom_point(size = 0.1) + 
  scale_colour_viridis_c(name = "Probability") + 
  ggtitle("Trajectory/E7.5/EHF") + 
  facet_grid(source ~ stage) + 
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

Plot E7.75 EHF trajectory:

```{r}
ggplot(dat, aes(UMAP.1, UMAP.2, colour = trajectory_7.75_ehf)) + 
  geom_point(size = 0.1) + 
  scale_colour_viridis_c(name = "Probability") + 
  ggtitle("Trajectory/E7.75/EHF") + 
  facet_grid(source ~ stage) + 
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

Plot E8.0 EHF trajectory:

```{r}
ggplot(dat, aes(UMAP.1, UMAP.2, colour = trajectory_8.0_ehf)) + 
  geom_point(size = 0.1) + 
  scale_colour_viridis_c(name = "Probability") + 
  ggtitle("Trajectory/E8.0/LHF") + 
  facet_grid(source ~ stage) + 
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

Plot E7.5 LHF trajectory:

```{r}
ggplot(dat, aes(UMAP.1, UMAP.2, colour = trajectory_7.5_lhf)) + 
  geom_point(size = 0.1) + 
  scale_colour_viridis_c(name = "Probability") + 
  ggtitle("Trajectory/E7.5/LHF") + 
  facet_grid(source ~ stage) + 
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

Plot E7.75 LHF trajectory:

```{r}
ggplot(dat, aes(UMAP.1, UMAP.2, colour = trajectory_7.75_lhf)) + 
  geom_point(size = 0.1) + 
  scale_colour_viridis_c(name = "Probability") + 
  ggtitle("Trajectory/E7.75/LHF") + 
  facet_grid(source ~ stage) + 
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

Plot E8.0 LHF trajectory:

```{r}
ggplot(dat, aes(UMAP.1, UMAP.2, colour = trajectory_8.0_lhf)) + 
  geom_point(size = 0.1) + 
  scale_colour_viridis_c(name = "Probability") + 
  ggtitle("Trajectory/E8.0/LHF") + 
  facet_grid(source ~ stage) + 
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

## Fate

Create a per-cell data frame:

```{r}
dat <- makePerCellDF(sce)
```

Plot E7.25 EHF fate:

```{r}
ggplot(dat, aes(UMAP.1, UMAP.2, colour = fates_7.25_ehf)) + 
  geom_point(size = 0.1) + 
  scale_colour_viridis_c(name = "Probability") + 
  ggtitle("Fate/E7.25/EHF") + 
  facet_grid(source ~ stage) + 
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

Plot E7.5 EHF fate:

```{r}
ggplot(dat, aes(UMAP.1, UMAP.2, colour = fates_7.5_ehf)) + 
  geom_point(size = 0.1) + 
  scale_colour_viridis_c(name = "Probability") + 
  ggtitle("Fate/E7.5/EHF") + 
  facet_grid(source ~ stage) + 
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

Plot E7.75 EHF fate:

```{r}
ggplot(dat, aes(UMAP.1, UMAP.2, colour = fates_7.75_ehf)) + 
  geom_point(size = 0.1) + 
  scale_colour_viridis_c(name = "Probability") + 
  ggtitle("Fate/E7.75/EHF") + 
  facet_grid(source ~ stage) + 
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

Plot E8.0 EHF fate:

```{r}
ggplot(dat, aes(UMAP.1, UMAP.2, colour = fates_8.0_ehf)) + 
  geom_point(size = 0.1) + 
  scale_colour_viridis_c(name = "Probability") + 
  ggtitle("Fate/E8.0/LHF") + 
  facet_grid(source ~ stage) + 
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

Plot E7.5 LHF fate:

```{r}
ggplot(dat, aes(UMAP.1, UMAP.2, colour = fates_7.5_lhf)) + 
  geom_point(size = 0.1) + 
  scale_colour_viridis_c(name = "Probability") + 
  ggtitle("Fate/E7.5/LHF") + 
  facet_grid(source ~ stage) + 
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

Plot E7.75 LHF fate:

```{r}
ggplot(dat, aes(UMAP.1, UMAP.2, colour = fates_7.75_lhf)) + 
  geom_point(size = 0.1) + 
  scale_colour_viridis_c(name = "Probability") + 
  ggtitle("Fate/E7.75/LHF") + 
  facet_grid(source ~ stage) + 
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

Plot E8.0 LHF fate:

```{r}
ggplot(dat, aes(UMAP.1, UMAP.2, colour = fates_8.0_lhf)) + 
  geom_point(size = 0.1) + 
  scale_colour_viridis_c(name = "Probability") + 
  ggtitle("Fate/E8.0/LHF)") + 
  facet_grid(source ~ stage) + 
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

## Selection

Take a random sample of 500 cells based on their trajectory probability:

```{r}
set.seed(0001001011)

N <- 500

itr <- c(
  "fates_7.25_ehf", 
  "fates_7.5_ehf", 
  "fates_7.75_ehf", 
  "fates_8.0_ehf", 
  "fates_7.5_lhf", 
  "fates_7.75_lhf", 
  "fates_8.0_lhf"
)

out <- lapply(itr, function(i) {
  
  p <- colData(sce)[, i]
  
  p[is.na(p)] <- 0
  
  sample(colnames(sce), N, replace = TRUE, prob = p)

})

names(out) <- itr

ids <- unlist(out)

ids <- c(ids, colnames(sce)[sce$sample %in% c("EHF", "LHF")])

ids <- unique(ids)
```

Subset experiment object by include annotation:

```{r}
alt <- sce[, ids]
```

Subset stage colours for downstream plotting:

```{r}
ind <- names(EmbryoStageColours) %in% alt$stage

EmbryoStageColours <- EmbryoStageColours[ind]
```

Subset theiler colours for downstream plotting:

```{r}
ind <- names(EmbryoTheilerColours) %in% alt$theiler

EmbryoTheilerColours <- EmbryoTheilerColours[ind]
```

Subset celltype colours for downstream plotting:

```{r}
ind <- names(EmbryoCelltypeColours) %in% alt$celltype

EmbryoCelltypeColours <- EmbryoCelltypeColours[ind]
```

## Correction

Create merge order for batch correction:

```{r}
dat <- table(alt$sample, alt$stage)

dat <- as.data.frame(dat, stringsAsFactors = FALSE)

colnames(dat) <- c("sample", "stage", "ncells")

dat <- subset(dat, ncells > 0)

dat <- subset(dat, !sample %in% c("EHF", "LHF"))

var <- c("E6.5", "E6.75", "E7.0", "mixed_gastrulation", "E7.25", "E7.5", "E7.75", "E8.0", "E8.25", "E8.5")

dat$stage <- factor(dat$stage, levels = var, ordered = TRUE)

dat <- dat[order(dat$stage, dat$ncells, decreasing = TRUE), ]

idx <- c(dat$sample, "LHF", "EHF")
```

Quickly perform batch correction:

```{r}
mnn <- quickCorrect(
  alt, 
  batch = alt$sample, 
  correct.all = TRUE, 
  PARAM = FastMnnParam(merge.order = idx),
  hvg.args = list(
    var.field = "bio",
    var.threshold = 0
  )
)

metadata(mnn$corrected)$combineVar <- mnn$dec

metadata(mnn$corrected)$getTopHVGs <- mnn$hvgs

mnn <- mnn$corrected
```

Copy column and assay data:

```{r}
colData(mnn) <- colData(alt)

logcounts(mnn) <- logcounts(alt)
```

## Dignostics

Perform UMAP on corrected PCA matrix:

```{r}
set.seed(1001001000)

mnn <- runUMAP(mnn, dimred = "corrected")
```

Create a per-cell data frame:

```{r}
dat <- makePerCellDF(mnn)

dat <- split(dat, dat$data)
```

Plot cells coloured by stage annotation:

```{r}
ggplot(dat$reference, aes(UMAP.1, UMAP.2, colour = stage)) + 
  geom_point(size = 0.1) + 
  scale_colour_manual(name = "Stage", values = EmbryoStageColours) + 
  guides(colour = guide_legend(override.aes = list(size = 2))) + 
  geom_point(data = dat$query, aes(UMAP.1, UMAP.2, shape = sample), inherit.aes = FALSE) + 
  scale_shape_manual(name = "Sample", values = c(1, 4)) + 
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

Plot cells coloured by theiler annotation:

```{r}
ggplot(dat$reference, aes(UMAP.1, UMAP.2, colour = theiler)) + 
  geom_point(size = 0.1) + 
  scale_colour_manual(name = "Theiler", values = EmbryoTheilerColours) + 
  guides(colour = guide_legend(override.aes = list(size = 2))) + 
  geom_point(data = dat$query, aes(UMAP.1, UMAP.2, shape = sample), inherit.aes = FALSE) + 
  scale_shape_manual(name = "Sample", values = c(1, 4)) + 
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

Plot cells coloured by celltype annotation:

```{r}
plt <- ggplot(dat$reference, aes(UMAP.1, UMAP.2, colour = celltype)) + 
  geom_point(size = 0.1) + 
  scale_colour_manual(name = "Celltype", values = EmbryoCelltypeColours) + 
  geom_point(data = dat$query, aes(UMAP.1, UMAP.2, shape = sample), inherit.aes = FALSE) + 
  scale_shape_manual(name = "Sample", values = c(1, 4)) + 
  theme_no_axes() + 
  theme(aspect.ratio = 1)

resizeGuides(plt, pointSize = 1, textSize = 7, spaceLegend = 0.6)
```

## Trajectory {.tabset}

Perform lineage inference with Slingshot:

```{r}
mnn <- slingshot(
  data = mnn, 
  reducedDim = "corrected", 
  dist.fun = dist_clusters_diag
)
```

Perform DiffusionMap on corrected PCA matrix:

```{r}
set.seed(1011000110)

mnn <- runDiffusionMap(mnn, dimred = "corrected")
```

### Lineage 1

Embed and construct curve data:

```{r}
sc1 <- embedCurves(mnn, "DiffusionMap")

sc1 <- slingCurves(sc1)[[1]]

sc1 <- data.frame(sc1$s[sc1$ord, ])

colnames(sc1) <- c("DiffusionMap.1", "DiffusionMap.2")
```

Create a per-cell data frame:

```{r}
dat <- makePerCellDF(mnn)

dat <- split(dat, dat$data)
```

Plot cells coloured by pseudotime annotation:

```{r}
ggplot(dat$reference, aes(DiffusionMap.1, DiffusionMap.2, colour = slingPseudotime_1)) + 
  geom_point(size = 0.1) + 
  scale_colour_viridis_c(name = "Pseudotime") + 
  geom_point(data = dat$query, aes(DiffusionMap.1, DiffusionMap.2, shape = sample), inherit.aes = FALSE) + 
  scale_shape_manual(name = "Sample", values = c(1, 4)) + 
  geom_path(data = sc1, aes(DiffusionMap.1, DiffusionMap.2), size = 1, arrow = arrow(), inherit.aes = FALSE) + 
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

Plot cells coloured by stage annotation:

```{r}
ggplot(dat$reference, aes(DiffusionMap.1, DiffusionMap.2, colour = stage)) + 
  geom_point(size = 0.1) + 
  scale_colour_manual(name = "Stage", values = EmbryoStageColours) + 
  guides(colour = guide_legend(override.aes = list(size = 2))) + 
  geom_point(data = dat$query, aes(DiffusionMap.1, DiffusionMap.2, shape = sample), inherit.aes = FALSE) + 
  scale_shape_manual(name = "Sample", values = c(1, 4)) + 
  geom_path(data = sc1, aes(DiffusionMap.1, DiffusionMap.2), size = 1, arrow = arrow(), inherit.aes = FALSE) + 
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

Plot cells coloured by theiler annotation:

```{r}
ggplot(dat$reference, aes(DiffusionMap.1, DiffusionMap.2, colour = theiler)) + 
  geom_point(size = 0.1) + 
  scale_colour_manual(name = "Theiler", values = EmbryoTheilerColours) + 
  guides(colour = guide_legend(override.aes = list(size = 2))) + 
  geom_point(data = dat$query, aes(DiffusionMap.1, DiffusionMap.2, shape = sample), inherit.aes = FALSE) + 
  scale_shape_manual(name = "Sample", values = c(1, 4)) + 
  geom_path(data = sc1, aes(DiffusionMap.1, DiffusionMap.2), size = 1, arrow = arrow(), inherit.aes = FALSE) + 
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

Plot cells coloured by celltype annotation:

```{r}
plt <- ggplot(dat$reference, aes(DiffusionMap.1, DiffusionMap.2, colour = celltype)) + 
  geom_point(size = 0.1) + 
  scale_colour_manual(name = "Celltype", values = EmbryoCelltypeColours) + 
  geom_point(data = dat$query, aes(DiffusionMap.1, DiffusionMap.2, shape = sample), inherit.aes = FALSE) + 
  scale_shape_manual(name = "Sample", values = c(1, 4)) + 
  geom_path(data = sc1, aes(DiffusionMap.1, DiffusionMap.2), size = 1, arrow = arrow(), inherit.aes = FALSE) + 
  theme_no_axes() + 
  theme(aspect.ratio = 1)

resizeGuides(plt, pointSize = 1, textSize = 7, spaceLegend = 0.6)
```

## Pseudotime {.tabset}

### Lineage 1

Test for differences along pseudotime:

```{r}
library(TSCAN)

res <- testPseudotime(
  x = logcounts(mnn),
  pseudotime = mnn$slingPseudotime_1,
  block = mnn$batch
)
```

Subset results by significance threshold:

```{r}
sig <- subset(res, FDR < 0.05)
```

Display table of significant results:

```{r}
knitr::knit_print(DT::datatable(as.data.frame(sig)))
```

Plot top 50 up-regulated genes:

```{r fig.width = 14, fig.height = 9}
sig.up <- subset(sig, logFC > 0)

sig.up <- sig.up[order(sig.up$logFC, decreasing = TRUE), ]

plotHeatmap(
  object = mnn, 
  features = head(rownames(sig.up), n = 50), 
  exprs_values = "reconstructed", 
  center = TRUE,
  colour_columns_by = c("celltype", "source", "stage"), 
  order_columns_by = "slingPseudotime_1", 
  column_annotation_colors = list(
    celltype = EmbryoCelltypeColours,
    source = EmbryoSourceColours,
    stage = EmbryoStageColours
  )
)
```

Plot top 50 down-regulated genes:

```{r fig.width = 14, fig.height = 9}
sig.down <- subset(sig, logFC < 0)

sig.down <- sig.down[order(sig.down$logFC, decreasing = FALSE), ]

plotHeatmap(
  object = mnn, 
  features = head(rownames(sig.down), n = 50), 
  exprs_values = "reconstructed", 
  center = TRUE,
  colour_columns_by = c("celltype", "source", "stage"), 
  order_columns_by = "slingPseudotime_1", 
  column_annotation_colors = list(
    celltype = EmbryoCelltypeColours,
    source = EmbryoSourceColours,
    stage = EmbryoStageColours
  )
)
```

## Summary

### Session

Print session information:

```{r}
sessionInfo()
```
